Certified Knowledge Compilation \\ with Application to Formally Verified Model Counting

### 2024-01-05 doc/JAIR/jair-paper.tex
Split Experimental results as separate file

\section{Experimental Evaluation}
\label{sect:experimental}

%%\input{figure-combined-verify}

\begin{figure}
\begin{minipage}{0.49\textwidth}
\centering{%
\begin{tikzpicture}[scale = 0.50]
  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log,
      legend style={at={(0.35,0.96)}},
      legend cell align={left},
                              x post scale=1.8, y post scale=2.4,
                              xmode=log,xmin=0.01,xmax=4000,
                              xtick={0.01, 0.1,1.0,10,100,1000,10000}, xticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}, {10,000}},
                              ymin=0.01, ymax=29000,
                              ytick={0.01, 0.1,1.0,10,100,1000,10000,100000}, yticklabels={0.01, 0.1, 1.0, 10, 100, {1,000},{10,000},{100,000}},
                              xlabel={D4 runtime (seconds)},
                              %ylabel={CPOG generation and checking runtime (seconds)},
%                              title={D4 Defining Clause Generation}
            ]
%    \input{data-formatted/time-d4-verify}

\addplot [only marks, color=redorange, mark=*,  mark options={scale=0.8}, opacity=0.7] coordinates { (5.758,13.334) (0.017,0.022) (0.014,0.034) (0.179,1.754) (0.016,0.046) (0.012,0.012) (0.024,0.13) (0.928,18.877) (0.069,0.624) (0.166,18.378) (1.48,21.674999999999997) (3.356,116.882) (3.184,46.001000000000005) (1.676,46.439) (0.016,0.063) (0.019,0.11100000000000002) (0.472,6.808) (0.02,0.126) (5.048,182.556) (0.398,74.411) (0.685,21.603) (0.252,3.414) (30.753,2292.7690000000002) (0.643,9.902) (2.154,36.441) (1.158,4.7219999999999995) (2.436,83.753) (1.636,132.504) (1.491,59.292) (4.683,1499.441) (30.566,4136.219) (145.967,2177.9669999999996) (2.666,40.712) (0.105,1.923) (5.508,31.97) (244.952,7273.887000000001) (0.206,8.503) (606.378,13198.325) (23.234,39.160000000000004) (0.837,182.089) (49.574,52.467) (25.077,42.550000000000004) (63.691,12374.461) (4.544,147.119) (1.679,359.63) (22.322,4493.414000000001) (2.076,61.509) (70.355,1322.4) (90.915,2908.19) (123.037,6910.268) (112.248,9602.115) (178.14,4147.301) (394.71,517.646) (158.115,180.629) (69.122,315.91) (508.972,258.496) (2.033,84.352) (270.954,433.669) (1536.914,859.557) (19.013,8856.852) (2.938,272.899) (598.709,8218.087) (9.311,105.213) (0.277,2.226) (0.101,1.086) (30.577,394.01700000000005) (0.18,2.051) (0.016,0.064) (0.021,0.134) (0.52,14.541) (0.016,0.063) (0.237,11.949000000000002) (0.985,58.088) (2.278,33.982) (1.053,16.815) (1.326,20.087) (2.908,45.419) (3.441,45.275) (2.06,30.061999999999998) (1.483,108.72) (3.377,47.61) (0.017,0.073) (0.016,0.052000000000000005) (3.536,72.871) (1.337,19.372999999999998) (5.587,81.767) (193.799,9953.933) (89.858,3959.588) (0.023,0.06899999999999999) (49.796,2973.447) (0.718,76.449) (5.364,204.765) (3.21,138.828) (41.455,1955.3980000000001) (1.083,53.278999999999996) (1.695,279.8) (1.345,286.818) (3.111,411.545) (6.397,1495.882) (22.214,666.7180000000001) (65.322,110.378) (22.917,420.612) (272.642,3116.728) (1.929,85.946) (1358.165,5872.383) (3951.002,10810.879) (14.808,1124.911) (93.537,4307.451)};

%    \input{data-formatted/time-d4-verify-onesided-only}

\addplot [only marks, color=bluegray, mark=square*,  mark options={scale=0.8}, opacity=0.7] coordinates { (205.743,2420.6859999999997) (326.575,4433.625) (476.021,3973.2659999999996) (494.38,5879.952) (1046.603,10779.982) (351.999,2069.865) (64.21,479.92100000000005) (316.505,3074.084) (788.377,6710.351000000001)};

%    \input{data-formatted/time-d4-failure}

\addplot [only marks, color=darkgreen, mark=diamond*, mark options={scale=1.1}, opacity=0.7] coordinates { (2052.239,25000) (424.73,25000) (2636.562,25000) (1403.431,25000) (3070.214,25000) (2053.446,25000) (2790.169,25000)};

    \legend{
 \textsf{Full validation},
 \textsf{One-sided validation},
 \textsf{Resource limit exceeded},
    }
    \addplot[mark=none, dashed] coordinates{(0.01,25000) (4000, 25000)};
    \addplot[mark=none, color=lightblue] coordinates{(0.01,20000) (4000,20000)};
    \addplot[mark=none, color=lightblue] coordinates{(0.1,0.01) (10000.0,1000.0)};
    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.01) (10000.0,10000.0)};
    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.1) (2000,20000)};
    \addplot[mark=none, color=lightblue] coordinates{(0.01,1.0) (200, 20000)};
    \addplot[mark=none, color=lightblue] coordinates{(0.01,10.0) (20, 20000)};
    \node[left] at (axis cs: 0.9,0.12) {$0{.}1\times$};
    \node[left] at (axis cs: 0.09,0.12) {$1\times$};
    \node[left] at (axis cs: 0.09,1.2) {$10\times$};
    \node[left] at (axis cs: 0.09,12.0) {$100\times$};
    \node[left] at (axis cs: 0.09,120.0) {$1000\times$};

          \end{axis}
\end{tikzpicture}
} % centering
\end{minipage}
\begin{minipage}{0.49\textwidth}
\centering{%
\begin{tikzpicture}[scale = 0.50]
  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10},
                              x post scale=1.8, y post scale=2.15,
                              xmode=log,xmin=100,xmax=1000000000, 
                              xtick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000}, xticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$},
                              ymode=log, ymin=100, ymax=1000000000, 
                              ytick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000}, yticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$},
                              xlabel={Defining Clauses},
                              %ylabel={Proof Clauses},
%                              title={D4 Defining Clause Generation}
            ]
    \addplot[mark=none, color=lightblue] coordinates{(100,100) (1e9,1e9)};
    \addplot[mark=none, color=lightblue] coordinates{(100,1000) (1e8,1e9)};
    \addplot[mark=none, color=lightblue] coordinates{(100,10000) (1e7,1e9)};
    \addplot[mark=none, color=lightblue] coordinates{(100,100000) (1e6,1e9)};
    \addplot[mark=none, color=lightblue] coordinates{(100,1000000) (1e5,1e9)};
%%    \input{data-formatted/defining+total}

\addplot [only marks, color=redorange, mark=*,  mark options={scale=0.8}, opacity=0.7] coordinates { (18271,161794) (1295,2397) (1659,4427) (127284,262373) (2325,6077) (359,554) (5373,14373) (314359,1213826) (44178,81780) (493,33207) (605988,1117440) (1141369,3809317) (1197649,2211944) (1001050,2393131) (2947,7851) (4991,13207) (355617,721935) (4897,13101) (2437497,5273728) (345,46079) (701346,1593947) (3191,26378) (5985623,28392373) (467550,940819) (1318024,2813565) (4713,23892) (1682059,3160386) (1406192,4043887) (899701,1704811) (4936014,11840667) (8038928,68469623) (70651620,197339867) (9567,251890) (39501,104182) (18007,187551) (125319627,374762287) (82194,297768) (269645146,770382773) (304,427342) (574062,8448281) (887,478450) (1757,399856) (78873354,503117290) (190248,429409) (2064844,13434107) (13737124,141587875) (50722,820648) (45923418,82243431) (83111464,188103550) (40776253,104351646) (79106742,265990284) (3133905,23973053) (16147,1697847) (45214,1277037) (72258,4612352) (905,2500003) (401032,840020) (355,2155900) (63050,4692737) (3195277,6098685) (1125105,4417869) (325251378,736764776) (6074672,13104909) (132891,287958) (65522,136116) (15561156,38336288) (143191,298755) (2851,7851) (5243,13911) (334862,681444) (2659,7335) (82585,317789) (447979,1612224) (905899,1733678) (433226,802268) (551970,1006319) (1891045,3387533) (1210098,2232426) (755273,1440924) (470934,1914084) (1178453,2188886) (3131,8575) (2515,6607) (1903735,4312279) (1107098,1868005) (1801878,3324418) (47556240,206213289) (24087642,83491057) (3109,8227) (16038150,101261480) (312728,642675) (3175308,7847036) (142059,747752) (20382514,69650789) (453769,1563728) (498246,1154507) (797866,6840229) (3835114,17412807) (5035130,10668725) (647011,9905453) (5189,1234644) (740235,2672407) (219979530,417455256) (402460,844395) (2295530,81147729) (1772352,119113610) (3975194,16763326) (8493275,41873955)};

    \node[left] at (axis cs: 900,1200) {$1\times$};
    \node[left] at (axis cs: 900,12000) {$10\times$};
    \node[left] at (axis cs: 900,120000) {$100\times$};
    \node[left] at (axis cs: 900,1200000) {$1{,}000\times$};
    \node[left] at (axis cs: 900,12000000) {$10{,}000\times$};

          \end{axis}
\end{tikzpicture}
} % centering
\end{minipage}
\caption{Runtime (left) and proof size (right) for CPOG proofs.  The runtime includes proof generation, checking, and model counting relative to the runtime for D4.
  The proof size is measured as total clauses relative to the number of defining clauses.}
\label{fig:dual:total}
\end{figure}

For our
evaluation, we used the public benchmark problems from the 2022
standard and weighted model competitions.\footnote{Downloaded from \url{https://mccompetition.org/2022/mc\_description.html}}  We found that there were
180 unique CNF files among these, ranging in size from 250 to
2,753,207 clauses.
We ran our programs on a processor with 64~GB of
memory and having an attached high-speed, solid-state
disk.
With a runtime limit of 4000 seconds, \dfour{} completed for 124 of the
benchmark problems.  Our proof generator was able to convert all of
these into POGs, with their declarations ranging from 304 to
2,761,457,765 (median 848,784) defining clauses.
%(The
%maximum count overflowed the overflowed the 32-bit signed integers we
%used to represent clause identifiers in both the proof generator and
%checker.)

We ran our proof generator with a time limit of 10,000 seconds.  The
results are shown in Figure~\ref{fig:dual:total}, The left-hand plot
shows the elapsed time for the combination of proof generation,
checking, and counting versus the time for
\dfour{}.
The proof generator was able to generate full proofs for 108 of the problems and
one-sided proofs for an additional 9 of them, leaving just 7 with no
verification.  The prototype checker successfully verified all of the generated
proofs.  The longest runtime for the combination of proof generation, checking, and counting
for a full proof was 13,145 seconds.
Overall the ratio
between the combined time for generation, checking, and counting versus the time
for \dfour{} had a harmonic mean of 5.5.
The right-hand plot shows the total number of clauses in the CPOG file versus the number of defining clauses for the problems having full proofs.
The ratio between the total number of clauses and the number of defining clauses had a harmonic mean of 3.13.
To date, we have not found
any errors in the dec-DNNF graphs generated by \dfour{}.

We found that the monolithic approach for generating the forward
implication proof works well for smaller POGs (up to one million
defining clauses), but it becomes inefficient for larger ones.  These
experiments suggest a possible hybrid approach, stopping the recursion
of the structural approach and shifting to monolithic mode once the
subgraph size is below some threshold.  By using monolithic mode, we
were also able to perform end-to-end verification of all but one of the
benchmarks that could be verified without preprocessing, with a total
time limit (including preprocessing) of 1,000 seconds.

We also found that our two optimizations: literal grouping and
lemmas can provide substantial improvements in proof size and runtime.
In the extreme cases, a lack
of lemmas caused one proof to grow by a factor of 52.5, while a lack
of literal grouping caused another proof to grow by a factor of 39.6.
Overall, it is clear that these two optimizations are worthwhile, and
sometimes critical, to success for some benchmarks.

Running the verified proof checker in \lean{} required, on average
(harmonic mean), around 5.9 times longer than the prototype checker.
Encouragingly, the scaling trends were identical for the two solvers,
indicating that the two checkers have similar asymptotic performance.
We consider a factor of 5.9 to be acceptable for the assurance
provided by formal verification.

In comparing other proof frameworks, we found that the \cdfour{}
toolchain ran very fast and could handle very large benchmarks.  Even
with a total time limit of 1,000 seconds, including the time for
knowledge compilation, the \cdfour{} toolchain completed 106
benchmarks, while the CPOG toolchain completed just 82.  We found
that the runtimes for the MICE toolchain versus our CPOG toolchain
showed little correlation, reflecting the fact that the two solve
different problems and use different approaches.  In general, our CPOG
toolchain showed better scaling, in part due to its ability to control
the recursion through lemmas.  Neither of these prior
toolchains could perform end-to-end verification when the knowledge
compilation was preceded by preprocessing.

>>>>

\input{experimental-section}
